The Spatial and Temporal Domains of Modern Ecology

### 2017-06-28 paper/main.tex
Final re-submission version of manuscript. Revised package structure so that input data will install as part of R package.  Corrected mistake in one perturbation parameter (for extent, which had uncertainty value slightly too low), re-rain analyses, updated figures in text (no substantive change).

\noindent \textbf{Results}\\
\noindent \textbf{\emph{Domains}}\\
In terms of resolution, the majority (67\%) of observations were collected in plots of $<$1 m$^2$ resolution, 24\% were collected within plots of 1 m$^2$ up to 1 ha, and the remaining 9\% in plots of $\geq$1 ha (Fig. 1A). The extent of 19\% of observations was $<$10 ha, 23\% covered 10-1,000 ha, 11\% 1,000-10,000 ha, 19\% 10,000-100,000 ha, 12\% 100,000-1,000,000 ha, and 16\% $>$1,000,000 ha (Fig. 1B).

@@

In the temporal dimensions, 37\% of observations were not repeated (Fig. 1C), 17\% were repeated at short intervals (sub-second to daily), 20\% at daily to monthly intervals, 18\% at monthly to yearly intervals, and 8\% at yearly to decadal intervals. Duration was one day or less for 31\% of sampled observations, while 10\% covered one day to one month, 23\% lasted one month to one year, 27\% covered 1-10 years, and 9\% spanned a decade or more (including several paleoecological studies covering centuries to millennia; Fig. 1D).

@@

To provide further insight into observational domains, we also evaluated the degree to which the scales \emph{represented} by observations differed from those that were actually observed. To make this assessment, we first log$_{10}$ transformed and then subtracted the values of i) actual extent from extent and ii) actual duration from duration, in order to calculate the magnitude of difference (or decade) between each pair of dimensions for each observation. We then compared how these magnitudes varied in relation to the scale of the actual dimension (Fig. 3). This comparison showed that most (81\%) of the assessed observations had actual extents of $\leq$1 ha, which on average were 4 to nearly 8 orders of magnitude smaller than the quantified extent (Fig 3.A).  Actual extent converged with extent $\geq$100,000 ha, but this applies to just 5\% of observations. The actual duration of 64\% of observations was $<$ 1 day, which on average was 3 to nearly 9 orders of magnitude shorter than the time span covered by temporal replicates (Fig. 3B). The two duration measures were approximately equal for the 16\% of observations exceeding one month of actual duration.

\begin{figure}[ht]
\includegraphics[width=1\textwidth]{figures/act_v_eff_diff.png}
\vspace{-0.2 cm}
\caption{The difference between \emph{actual} extent (the summed area of spatial replicates) and extent (A) and \emph{actual} duration (the summed sampling duration across temporal replicates) and duration (B). Differences are expressed as decades, or how many orders of magnitude greater the extent or duration is then the actual extent or duration, and are summarized (as box plots, with circle in box representing the mean and line the median) in bins representing different levels of actual extent/duration.  The percentages of observations falling within each bin are indicated by the color of the inter-quartile and the numeric value above the upper whisker.}
\label{afoto1}
\end{figure}

@@

Have these domains changed since the seminal papers on scale first began to appear in the late 1980s \cite{wiens_spatial_1989, levin_problem_1992, tilman_ecological_1989}? A comprehensive answer to this question would require a similar study focused on earlier literature, but an analysis of results presented in three prior studies provides partial insight. The first and most comprehensive dataset consists of duration values extracted by Tilman \cite{tilman_ecological_1989} from 623 studies published between 1977-1987 in the journal \emph{Ecology}. The average duration of the most comparable subset of those values (n=419; see SI) was 3.6 years, compared to 3.3 years for observations in our sample (or 5.1 if temporally un-replicated observations are excluded). The second dataset is found in Kareiva and Anderson \cite{kareiva_spatial_1988}, who present the resolutions of 97 community ecology experiments published in \emph{Ecology} between 1980-1986. The average of those (12,657 m$^2$) was substantially smaller than the mean of our sample (1,510,247 m$^2$), but comparing the 80th percentile value (197 m$^2$) of Kareiva and Anderson's \cite{kareiva_spatial_1988} to ours (115 m$^2$) shows that the majority of contemporary observations are finer-grained than most 1980s-era experiments. The third dataset is provided by Porter et al \cite{porter_wireless_2005}, who compared the extent and interval of 25 studies published in 2003 and 2004 (also in \emph{Ecology}). The mean interval was 178 days, compared to 684 days in our sample, but the 80th percentile value in our study was 169 days compared to 329 days in theirs. Extent in our sample was substantially larger according to multiple summary statistics, including the mean (115,098,283 ha in our study versus 368,403 ha), median (5,120 ha versus 9 ha), and 95th percentile (45,104,594 ha versus 136,000 ha).

@@

It is therefore important to examine whether the phenomena being observed are adequately captured by the design of replicates. Our methods suggest one possible procedure for assessing the representativeness of replicates: 1) measure auto-correlation (spatial or serial) in the replicates, 2) add the autocorrelation length to the replicate area/duration, 3) calculate an autocorrelation-adjusted actual extent/duration, and 4) plot where it falls between actual extent/duration and extent/duration. If the adjusted actual value is substantially closer to the representative value, then this may provide greater confidence in the adequacy of the sampling scheme. If not, then alternative sampling methods may be used to close this gap. For example, remote sensing provides wall-to-wall spatial coverage of a study area, erasing the difference between actual extent and extent. Furthermore, the interval of high-resolution imaging (higher resolution is preferred in images as it allows individual features to be better discerned \cite{dark_modifiable_2007, hay_comparison_2003}) is now approaching daily to sub-daily scales \cite{drusch_sentinel-2:_2012, hand_startup_2015}, allowing improved representation of spatial and temporal dynamics. For phenomena that can't be measured from space, either because they are not visible or because they require continuous observation, new approaches for collecting \emph{in situ} or near-surface observations (e.g. low-cost wireless sensors \cite{wolf_gsm-based_2012, collins_new_2006, porter_wireless_2005}, citizen observers \cite{dickinson_current_2012}, and autonomous vehicles \cite{anderson_lightweight_2013}) can be used to increase the spatial and temporal coverage of observations.

>>>>

\noindent \textbf{Results}\\
\noindent \textbf{\emph{Domains}}\\
In terms of resolution, the majority (67\%) of observations were collected in plots of $<$1 m$^2$ resolution, 24\% were collected within plots of 1 m$^2$ up to 1 ha, and the remaining 9\% in plots of $\geq$1 ha (Fig. 1A). The extent of 19\% of observations was $<$10 ha, 23\% covered 10-1,000 ha, 12\% 1,000-10,000 ha, 19\% 10,000-100,000 ha, 12\% 100,000-1,000,000 ha, and 15\% $>$1,000,000 ha (Fig. 1B).

@@

In the temporal dimensions, 37\% of observations were not repeated (Fig. 1C), 17\% were repeated at short intervals (sub-second to daily), 20\% at daily to monthly intervals, 18\% at monthly to yearly intervals, 6\% at yearly to decadal intervals, and 2\% at decadal or greater intervals. Duration was one day or less for 31\% of sampled observations, while 10\% covered one day to one month, 23\% lasted one month to one year, 27\% covered 1-10 years, and 9\% spanned a decade or more (including several paleoecological studies covering centuries to millennia; Fig. 1D).

@@

For our final analysis of observational domains, we evaluated the differences between the scales represented by the extent and duration dimensions and the scales that ecological observations actually cover. To make this assessment, we log$_{10}$ transformed and then subtracted the values of i) actual extent from extent and ii) actual duration from duration, yielding the magnitude of difference between each pair of dimensions for each observation. We plotted these values (summarized in box plots) against their corresponding actual extent/duration values to evaluate whether these differences vary with scale (Fig. 3). This analysis showed that most (81\%) of the assessed observations had actual extents of $\leq$1 ha, which on average were 4 to nearly 8 orders of magnitude smaller than the primary measure of extent (Fig 3A). These differences declined linearily with actual extent, becoming approximately equal for the 5\% of observations having $\geq$100,000 ha actual extent. The actual duration of 64\% of observations was $<$1 day, which on average was 3 to nearly 9 orders of magnitude shorter than the time span covered by the conventional metric of duration (Fig. 3B). As with extent, the magnitude of difference declined steadily with actual duration, effectively disappearing for the 16\% of observations with $\geq$1 month of actual duration.

\begin{figure}[ht]
\includegraphics[width=1\textwidth]{figures/act_v_eff_diff.png}
\vspace{-0.2 cm}
\caption{The difference between extent and \emph{actual} extent (the summed area of spatial replicates) (A) and duration and \emph{actual} duration (the summed sampling duration across temporal replicates) (B). Difference values are expressed in terms of how many orders of magnitude larger (or longer) extent (duration) is than actual extent (actual duration), and are summarized (as box plots, with circle in box representing the mean and line the median) in bins representing increasing scales of actual extent/duration.  The percentages of observations falling within each bin are indicated by the color of the inter-quartile and the numeric value above the upper whisker.}
\label{afoto1}
\end{figure}

@@

Have these domains changed since the seminal papers on scale first began to appear in the late 1980s \cite{wiens_spatial_1989, levin_problem_1992, tilman_ecological_1989}? A comprehensive answer to this question would require a similar study focused on earlier literature, but an analysis of results presented in three prior studies provides partial insight. The first and most comprehensive dataset consists of duration values extracted by Tilman \cite{tilman_ecological_1989} from 623 studies published between 1977-1987 in the journal \emph{Ecology}. The average duration of the most comparable subset of those values (n=419; see SI) was 3.6 years, compared to 3.3 years for observations in our sample (or 5.1 if temporally un-replicated observations are excluded). The second dataset is found in Kareiva and Anderson \cite{kareiva_spatial_1988}, who present the resolutions of 97 community ecology experiments published in \emph{Ecology} between 1980-1986. The average of those (12,657 m$^2$) was substantially smaller than the mean of our sample (1,496,069 m$^2$), but comparing the 80th percentile value (197 m$^2$) of Kareiva and Anderson's \cite{kareiva_spatial_1988} to ours (115 m$^2$) shows that the majority of contemporary observations are finer-grained than most 1980s-era experiments. The third dataset is provided by Porter et al \cite{porter_wireless_2005}, who compared the extent and interval of 25 studies published in 2003 and 2004 (also in \emph{Ecology}). The mean interval was 178 days, compared to 684 days in our sample, but the 80th percentile value in our study was 169 days compared to 329 days in theirs. Extent in our sample was substantially larger according to multiple summary statistics, including the mean (114,965,072 ha in our study versus 368,403 ha), median (5,051 ha versus 9 ha), and 95th percentile (46,424,808 ha versus 136,000 ha).

@@

It is therefore important to examine whether the scales of the phenomena being observed are adequately captured by the design of replicates. Our methods suggest one possible procedure for assessing the \emph{scale representativeness} of replicates: 1) measure auto-correlation (spatial or serial) in the replicates, 2) add the autocorrelation length to the replicate area/duration, 3) calculate an autocorrelation-adjusted actual extent/duration, and 4) plot where it falls between actual extent/duration and extent/duration. The distance between the adjusted actual value and the ostensible value can provide a measure of how well the replicates represent the intended scale of observation. If the gap remains large, then alternative sampling methods may be used to close it. For example, remote sensing provides wall-to-wall spatial coverage of a study area, erasing the difference between actual extent and extent. Furthermore, the interval of high-resolution imaging (higher resolution is preferred in images as it allows individual features to be better discerned \cite{dark_modifiable_2007, hay_comparison_2003}) is now approaching daily to sub-daily scales \cite{drusch_sentinel-2:_2012, hand_startup_2015}, allowing improved representation of spatial and temporal dynamics. For phenomena that can't be measured from space, either because they are not visible or because they require continuous observation, new approaches for collecting \emph{in situ} or near-surface observations (e.g. low-cost wireless sensors \cite{wolf_gsm-based_2012, collins_new_2006, porter_wireless_2005}, citizen observers \cite{dickinson_current_2012}, and autonomous vehicles \cite{anderson_lightweight_2013}) can be used to increase the spatial and temporal coverage of observations.
